A Machine Profile Summary

### 2022-01-18 MachineSummary/MachineSummary.Rmd
updated

If we want to compare results computed on different machines, we need a way to provide 
measures and identifiers of the particular computing environment at hand. This write-up
attempts to summarize some possibilities available in the R statistical software and language.

@@

@@

?? get_ram may not work in Windows 10.

## calceps.R

@@

```{r envrontest}
envron<-function(){
D1<-as.numeric(1) # use variable to avoid constants when double precision invoked
E5 <- 10 #arbitrary scaling for additive equality tests
B9 <- 1E+35 # big number, not necessarily biggest possible
E6 <- 1 # initial value for radix 
E9 <- 1 # initial value for machine precision
D0 <- E6 
repeat{
  E9 <- E9/2 #start of loop to decrease estimated machine precision
  D0 <- E6+E9 #force storage of sum into a floating-point scalar
  if (D0 <= E6) break
} # repeat reduction while (1+E9) > 1
E9 <- E9*2 # restore smallest E9 which gives (1+E9) > 1
repeat {
   E6 <- E6+1 # try different radix values
   D0 <- E6+E9
   if (D0 <= E6) break
} # until a shift is observed
J1 <- 1 # initial count of radix digits in mantissa
E9 <- 1 # use radix for exact machine precision
repeat {  
  E9 <- E9/E6 #loop while dividing by radix
  J1 <- J1+1 #increment counter
  D0 <- D1+E9 #add tp 1
  if (D0<=D1) break #test and repeat until equality
}
E9 <- E9*E6 #recover last value of machine precision
J1 <- J1-1 #and adjust the number of digits
mpvals<-list(eps=E9, radix=E6, ndigits=J1)
}
mp<-envron()
E9<-mp$eps
E6<-mp$radix
J1<-mp$ndigits
cat("E9 = ",E9,"  -- the machine precision,  E9=MIN (X  1+X>1)\n")
cat( "E9*1E+16=",E9*1E+16,"\n")
cat( "E6 = ",E6,"  -- the radix of arithmetic","\n")
cat( "J1 = ",J1,"  -- the number of radix digits in mantissa of FP numbers","\n")
cat( "E6^(-J1+1) * 1E+16=",1E+16*(E6^(-J1+1)),"\n")
cat(".Machine$double.eps=",.Machine$double.eps,"\n")
```

- calceps.R example

??JN exp(sin(cos())) benchmark

@@

From the above discussion, the following offers a possible compact solution.

@@

>>>>

If we want to compare results computed on different machines, we need a way to provide 
measures and identifiers of the particular computing environment at hand. This write-up
attempts to summarize some possibilities available in the R statistical software and language.
Note that one of our interests is to have tools that work in R **across different platforms**.
That is, we want to be able to call an R function to get machine and computing environment
characterizations regardless of the particular platform which we are currently using.
Moreover, we want to have a succinct output of such information.

@@

An important tool for acquiring machine information is the CRAN package
`benchmarkme` (@benchmarkme). We found a minor bug in the `get_ram()`
function of this package for some Windows 10 versions which is now 
corrected (??check it is uploaded). ??Do we wish to comment on the
"GB" issue?

@@

## Calculating the machine precision

@@

```{r envrontest}
envron<-function(){
D1<-as.numeric(1) # use variable to avoid constants when double precision invoked
E5 <- 10 #arbitrary scaling for additive equality tests
B9 <- 1E+35 # big number, not necessarily biggest possible
E6 <- 1 # initial value for radix 
E9 <- 1 # initial value for machine precision
D0 <- E6 
repeat{
  E9 <- E9/2 #start of loop to decrease estimated machine precision
  D0 <- E6+E9 #force storage of sum into a floating-point scalar
  if (D0 <= E6) break
} # repeat reduction while (1+E9) > 1
E9 <- E9*2 # restore smallest E9 which gives (1+E9) > 1
repeat {
   E6 <- E6+1 # try different radix values
   D0 <- E6+E9
   if (D0 <= E6) break
} # until a shift is observed
J1 <- 1 # initial count of radix digits in mantissa
E9 <- 1 # use radix for exact machine precision
repeat {  
  E9 <- E9/E6 #loop while dividing by radix
  J1 <- J1+1 #increment counter
  D0 <- D1+E9 #add tp 1
  if (D0<=D1) break #test and repeat until equality
}
E9 <- E9*E6 #recover last value of machine precision
J1 <- J1-1 #and adjust the number of digits
mpvals<-list(eps=E9, radix=E6, ndigits=J1)
}
mp<-envron()
E9<-mp$eps
E6<-mp$radix
J1<-mp$ndigits
cat("E9 = ",E9,"  -- the machine precision,  E9=MIN (X  1+X>1)\n")
cat( "E9*1E+16=",E9*1E+16,"\n")
cat( "E6 = ",E6,"  -- the radix of arithmetic","\n")
cat( "J1 = ",J1,"  -- the number of radix digits in mantissa of FP numbers","\n")
cat( "E6^(-J1+1) * 1E+16=",1E+16*(E6^(-J1+1)),"\n")
cat(".Machine$double.eps=",.Machine$double.eps,"\n")
cat("From R  .Machine:\n")
cat("double.eps=", .Machine$double.eps,"\n", 
    "double.base=", .Machine$double.base,"\n", 
    "double.digits=", .Machine$double.digits,"\n" )
```

## Simple performance examples

There are many performance examples used for testing the speed of computers. A 
well-known example is the @linbench03 LINPACK test used for ranking supercomputers.
These generally employ linear algebra calculations. The example below, by
contrast, focuses on the special function computations. Variations on this
simple performance test have been used by one of the authors for nearly half
a century. Note the use of the `microbenchmark` timing package (@microbench13).

```{r loopesc}
loopesc <- function(nn){
  ss<-0
  for (i in 1:nn) {
    xx<-exp(sin(cos(1.0*i)))
    ss<-ss+xx
  }
  xx
}
require("microbenchmark")
nvals<-c(1000, 10000, 100000, 1000000)
tvals<-rep(NA,length(nvals))
i<-0
for (nn in nvals) {
  i<-i+1
  cat(nn,"\n")
  tt<-microbenchmark(loopesc(nn), unit='us')
  print(tt)
  tvals[i]<-mean(tt$time)
}
plot(log10(nvals),log(tvals))

```

@@

From the above discussion, the following offers a possible compact solution. Users may
wish to modify this to their own particular needs.

@@

## References
